---
title: "Assignment 3: Verify the Claim"
subtitle: "Analyzing ACS data with R"
author: "Nathan Leroy"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

This analysis attempts to measure the accuracy of a 2018 article by the
Philadelphia Media Network, "[New Census figures on Philly neighborhoods
show inequality, high numbers of whites living in
poverty](https://whyy.org/articles/new-census-figures-on-philly-neighborhoods-show-inequality-high-numbers-of-whites-living-in-poverty/)".
This article uses American Community Survey (ACS) data to conduct a
neighborhood-by-neighborhood analysis of poverty and demographic
conditions across Philadelphia.

According to the article, conditions in Philadelphia have gotten
increasingly worse over the last 2 decades, with the following examples
as proof:

- Poverty rates in the neighborhood of Eastwick increased from 12 to 26
  percent between the 2006-10 census and the 2013-17 census.

- The median household income of Tioga-Nicetown declined from \$28,026
  to \$17,493.

- White poverty increased drastically throughout the city, with the
  neighborhood of Holmesburg showing an over 900 percent increase (2% to
  19%).

- The neighborhood of Fairhill had a poverty rate of 61 percent, and its
  median income dropped by \$300.

- Visualizations of changes in poverty rates in Philadelphia show a
  significant increase in many of the poorer areas of the city.

The purpose of this analysis is to verify some of the claims made in the
article by reconstructing the analysis of the ACS data and testing if
the reported changes in the article are statistically significant.

## Part 1: Import Data & Create Tables

First, we install all required packages. An API key from the US Census
Bureau is required for this project, and it needs to be installed before
any data can be fetched. A key can be easily obtained
[here](https://api.census.gov/data/key_signup.html).

```{r}
# Install required packages

library(tidycensus)
library(tidyverse)
```

After installing all the packages and confirming the API key is set up,
we can start importing data from the ACS table.

For this analysis, we will use the following codes:

- B17001_001: total population of an area for which poverty can be
  determined.

- B17001_002: total number of people in that area who fall below the
  poverty line.

```{r}
# Variables from ACS table: total in poverty and total below
poverty_vars <- c(
  total_poverty = "B17001_001",
  below_poverty = "B17001_002"
)
```

```{r}
# 2013–2017 ACS 5-year estimates

#| message: false
#| warning: false

poverty_2013_2017 <- get_acs(
  geography = "tract",
  variables = poverty_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2017,
  survey = "acs5",
  output = "wide",
  moe_level = 90
)

poverty_2013_2017 <- poverty_2013_2017 %>%
  mutate(
    period = "2013–2017",
    poverty_rate = (below_povertyE / total_povertyE) * 100
  )
```

```{r}
# 2006–2010 ACS 5-year estimates

#| message: false
#| warning: false

poverty_2006_2010 <- get_acs(
  geography = "tract",
  variables = poverty_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2010,
  survey = "acs5",
  output = "wide",
  moe_level = 90
)

poverty_2006_2010 <- poverty_2006_2010 %>%
  mutate(
    period = "2006–2010",
    poverty_rate = (below_povertyE / total_povertyE) * 100
  )
```

Here, we combine the data from each census into a single table for
easier analysis.

```{r}
poverty_both <- bind_rows(
  poverty_2006_2010,
  poverty_2013_2017
) %>%
  select(
    GEOID,
    NAME,
    period,
    total_povertyE,
    total_povertyM,
    below_povertyE,
    below_povertyM,
    poverty_rate
  ) %>%
  arrange(GEOID, period)

poverty_both
```

```{r}
poverty_change <- poverty_both %>%
  select(GEOID, NAME, period, poverty_rate) %>%
  pivot_wider(
    names_from = period,
    values_from = poverty_rate
  ) %>%
  mutate(
    change_percentage_points = `2013–2017` - `2006–2010`
  ) %>%
  arrange(desc(change_percentage_points))

# View the tracts with the largest increases
poverty_change %>%
  select(
    GEOID,
    NAME,
    `2006–2010`,
    `2013–2017`,
    change_percentage_points
  ) %>%
  print(n = 20)
```

## Part 2: Testing

In this portion of our analysis, we want to determine how reliable our
data for each census tract is. First, we calculate the MOE of the
poverty rate for each census.

```{r}
# Calculate poverty rate and its MOE

# 2006-2010
poverty_2006_2010 <- poverty_2006_2010 %>%
  mutate(
    poverty_rate = below_povertyE / total_povertyE,
    
    poverty_rate_moe = moe_prop(
      num = below_povertyE,
      denom = total_povertyE,
      moe_num = below_povertyM,
      moe_denom = total_povertyM
    )
  )

# 2013-2017
poverty_2013_2017 <- poverty_2013_2017 %>%
  mutate(
    poverty_rate = below_povertyE / total_povertyE,
    
    poverty_rate_moe = moe_prop(
      num = below_povertyE,
      denom = total_povertyE,
      moe_num = below_povertyM,
      moe_denom = total_povertyM
    )
  )
```

We also want to calculate the coefficient of variation. We can do this
using the two formulas below:

$$ SE = \frac{\text{MOE}}{1.645} $$ And:
$$ CV = \frac{\text{SE}}{estimate} * 100 $$

We use the thresholds determined by Jurjevich et al. (2018) for this
analysis:

- CV \< 12% is reliable

- CV \> 12% and CV \< 40% is somewhat reliable

- CV \> 40% is unreliable

```{r}
# CV Calculations

poverty_2006_2010 <- poverty_2006_2010 %>%
  mutate(
    poverty_rate_se = poverty_rate_moe / 1.645,
    poverty_rate_cv = (poverty_rate_se / poverty_rate) * 100,
    
    reliability = case_when(
      poverty_rate_cv < 12 ~ "Reliable",
      poverty_rate_cv <= 40 ~ "Somewhat reliable",
      poverty_rate_cv > 40 ~ "Unreliable",
      TRUE ~ NA_character_
    )
  )

poverty_2013_2017 <- poverty_2013_2017 %>%
  mutate(
    poverty_rate_se = poverty_rate_moe / 1.645,
    poverty_rate_cv = (poverty_rate_se / poverty_rate) * 100,
    
    reliability = case_when(
      poverty_rate_cv < 12 ~ "Reliable",
      poverty_rate_cv <= 40 ~ "Somewhat reliable",
      poverty_rate_cv > 40 ~ "Unreliable",
      TRUE ~ NA_character_
    )
  )
```

After these calculations, we join the two periods together to create one
dataset with one row per tract using an inner join.

::: callout-tip
We use an inner join because we want tracts that have data in both
periods for the change calculation.
:::

```{r}
poverty_change <- poverty_2006_2010 %>%
  select(
    GEOID,
    NAME,
    poverty_rate_2006_2010 = poverty_rate,
    poverty_moe_2006_2010 = poverty_rate_moe,
    poverty_cv_2006_2010 = poverty_rate_cv,
    reliability_2006_2010 = reliability
  ) %>%
  inner_join(
    poverty_2013_2017 %>%
      select(
        GEOID,
        poverty_rate_2013_2017 = poverty_rate,
        poverty_moe_2013_2017 = poverty_rate_moe,
        poverty_cv_2013_2017 = poverty_rate_cv,
        reliability_2013_2017 = reliability
      ),
    by = "GEOID"
  )
```

Finally, we calculate the change in poverty rates by simply subtracting
the 2006-10 rate from the newer 2013-17 rate.

```{r}
poverty_change <- poverty_change %>%
  mutate(
    change = poverty_rate_2013_2017 - poverty_rate_2006_2010
  )

# To make it easier to read, convert proportions into percents

poverty_change <- poverty_change %>%
  mutate(
    change_percentage_points = change * 100
  )
```

To calculate the MOE of the change between the two periods, we can use
this formula:

$$\text{MOE}_{\text{change}} = \sqrt{\text{MOE}_1^2 + \text{MOE}_2^2}$$

```{r}
poverty_change <- poverty_change %>% mutate(
  change_moe = sqrt( poverty_moe_2006_2010^2 + poverty_moe_2013_2017^2 ) )
```

```{r}
poverty_change <- poverty_change %>%
  mutate(
    change_lower = change - change_moe,
    change_upper = change + change_moe
  )
```

Now, we can check if the change is statistically significant at the 90%
confidence interval. If the change is larger than the MOE we calculated
earlier, the it is considered to be statistically significant.

```{r}
poverty_change <- poverty_change %>%
  mutate(
    change_significant = abs(change) > change_moe
  )
```

Several census tracts have missing or incomplete values. We can run the
script below to identify all of these tracts before deciding what to do
with them.

```{r}
poverty_change %>%
  filter(
    is.na(poverty_rate_2006_2010) |
      is.na(poverty_rate_2013_2017) |
      is.na(change)
  ) %>%
  select(
    GEOID,
    NAME,
    poverty_rate_2006_2010,
    poverty_rate_2013_2017,
    change
  )
```

For this analysis, we will drop any incomplete rows, as the poverty
level cannot be accurately calculated.

```{r}
poverty_analysis <- poverty_change %>%
  filter(
    !is.na(change),
    !is.na(change_moe),
    !is.na(change_significant)
  )
```

After the incomplete rows are dropped, we can calculate the total amount
of tracts with a detectable change in poverty level. 63/313
(approximately 16.75 percent) of the tracts show a statistically
significant change in poverty level.

```{r}
poverty_analysis %>%
  summarise(
    detectable_change = sum(change_significant),
    no_detectable_change = sum(!change_significant),
    total_tracts = n(),
    percent_detectable = mean(change_significant) * 100,
    percent_not_detectable = mean(!change_significant) * 100
  )
```

This table displays all the calculated data associated with the tracts.
The tracts with the greatest change in poverty level are displayed on
top.

```{r}
poverty_analysis %>%
  select(
    GEOID,
    NAME,
    poverty_rate_2006_2010,
    poverty_moe_2006_2010,
    poverty_rate_2013_2017,
    poverty_moe_2013_2017,
    change_percentage_points,
    change_moe,
    reliability_2006_2010,
    reliability_2013_2017,
    change_significant
  ) %>%
  arrange(desc(abs(change_percentage_points)))
```

## Part 3: Results and Visualizations

### Map

We haven't imported the geometry of the tracts yet, so we need to run
the following code block:

```{r}

#| include: false

library(sf)

invisible(capture.output(
  tracts <- get_acs(
    geography = "tract",
    variables = "B01003_001",
    state = "PA",
    county = "Philadelphia",
    year = 2017,
    survey = "acs5",
    geometry = TRUE
  )
))
```

Next, we join it to the main table.

```{r}
poverty_map <- tracts %>%
  select(GEOID, geometry) %>%
  left_join(
    poverty_analysis,
    by = "GEOID"
  )
```

Finally, we display the map.

```{r}
#| fig-height: 8
poverty_map %>%
mutate(
  change_percentage_points = change * 100
) %>%
  ggplot() +
  geom_sf(
    aes(fill = change_percentage_points),
    color = "black"
  ) +
  scale_fill_gradient2(
    name = "Change in poverty rate\n(percentage points)"
  ) +
  theme_void() +
  labs(
    title = "Change in Poverty Rates Across Philadelphia",
    subtitle = "Census tract ACS 5-year estimates, 2006–2010 vs. 2013–2017",
    caption = "Grey values indicate no data."
  )
```

This map displays where there is a statistically significant change in
poverty levels between the two censuses.

```{r}
#| fig-height: 7
poverty_map %>%
  mutate(
    change_status = case_when(
      change_significant & change > 0 ~ "Detectable increase",
      change_significant & change < 0 ~ "Detectable decrease",
      TRUE ~ "No detectable change"
    )
  ) %>%
  ggplot() +
  geom_sf(
    aes(fill = change_status),
    color = "black",
    linewidth = 0.3
  ) +
  theme_void() +
  labs(
    title = "Statistically Significant Changes in Philadelphia Poverty Rates",
    subtitle = "Census tract ACS 5-year estimates, 2006–2010 vs. 2013–2017",
    fill = "Change status",
    caption = "A detectable change falls outside of the 90% confidence interval."
  )
```

### Errorbar Visualization

This visualization shows only the statistically significant changes in
poverty rates between the two time periods. The bars represent the
estimated change, while the error bars (thinner black bars) represent
the 90% margin of error.

```{r}
poverty_analysis %>%
  filter(change_significant) %>%
  arrange(change) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(GEOID, change), y = change * 100)) +
  geom_col(aes(fill = change > 0)) +
  geom_errorbar(
    aes(
      ymin = (change - change_moe) * 100,
      ymax = (change + change_moe) * 100
    ),
    width = 0.3
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Statistically Significant Changes in Poverty Rates",
    subtitle = "Philadelphia census tracts, 2006–2010 vs. 2013–2017",
    x = "Census tract",
    y = "Change in poverty rate (percentage points)",
    caption = "Bars show estimated change; error bars show the 90% margin of error."
  )
```

### Reliability Table

This table displays the reliability of census tract poverty estimates
based on their CV. Lower CV values indicate more reliable estimates.

::: {.content-visible when-format="html"}
<details>

<summary><strong>Reliability table</strong></summary>

```{r}
library(gt)

reliability_table <- poverty_analysis %>%
  mutate(
    cv = pmax(
      poverty_cv_2006_2010,
      poverty_cv_2013_2017,
      na.rm = TRUE
    ),
    dot = case_when(
      cv < 12 ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE ~ "🔴"
    ),
    flag = case_when(
      cv < 12 ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE ~ "Unreliable"
    )
  ) %>%
  arrange(desc(cv)) %>%
  select(
    dot,
    GEOID,
    poverty_cv_2006_2010,
    poverty_cv_2013_2017,
    flag
  )

reliability_table %>%
  gt() %>%
  fmt_number(
    columns = c(poverty_cv_2006_2010, poverty_cv_2013_2017),
    
  ) %>%
  cols_label(
    dot = "",
    GEOID = "Census tract",
    poverty_cv_2006_2010 = "CV (%)\n2006–2010",
    poverty_cv_2013_2017 = "CV (%)\n2013–2017",
    flag = "Reliability"
  ) %>%
  cols_align( align = "center", columns = dot
  ) %>%
  tab_header(
    title = "Reliability of Philadelphia Census Tract Poverty Estimates",
    subtitle = "ACS 5-year estimates, 2006–2010 and 2013–2017"
  ) %>%
  tab_source_note(
    "Reliability thresholds follow Jurjevich et al. (2018): CV < 12% reliable, 12–40% somewhat reliable, > 40% unreliable."
  )
```

</details>
:::

## Part 4: Adjudication of Claims

A challenge of this project is that the census tracts found in the ACS
dataset do not correspond to traditional boundaries of Philadelphia
neighborhoods, which are somewhat informal and not collected or defined
by the city for official use. Therefore, we have to estimate which
tracts fall into the neighborhoods the article is analyzing. This
process is naturally a source of error in this project.

![Philadelphia Neighborhood Maps - Revolve Philly Group \| Map
...](https://i.pinimg.com/736x/16/09/2d/16092d197f53e945fcdd5f2ad911574e.jpg){width="506"}

The article particularly analyzes:

- Fairhill (26): located in North Philadelphia; a predominantly white
  neighborhood. This registered the highest poverty rate in the city, 61
  percent. This article explains how low-income neighborhoods are at
  particular risk of falling further into poverty, using Fairhill as an
  example.

  - Analysis: While the area around Fairhill did indeed see an average
    increase in poverty based on the map, few of these changes were
    statistically significant. Although the data does show that poverty
    increased slightly here, there is limited statistical evidence that
    poverty increased beyond the bounds of the variability of the ACS
    data.

- Eastwick (10): located Southwest Philadelphia: a predominantly black
  and low-income neighborhood. This neighborhood showed the greatest
  poverty-rate increase of any community, going from 12 to 26 percent.

  - Analysis: From our map in part 3, there did not appear to be any
    statistically significant increases in poverty rate during the study
    period. In fact, some of the areas showed a statistically
    significant decrease in poverty rate. This contrasts with the claim
    of the article that the poverty rate has more than doubled in this
    area. Since all of our census tract data from Southwest Philadelphia
    shows either statistically significant decline in poverty or no
    statistically significant change, we can confidently refute this
    claim.

- Holmesburg (50): located in Northeast Philadelphia: a predominantly
  white, working-class neighborhood. The article claims that poverty
  skyrocketed here, increasing more than 9x from 2 percent to 19
  percent.

  - In the Northeast Philadelphia area, there is a mix of statistically
    significant increases in poverty, statistically significant
    decreases, and non-significant changes. It seems unlikely for a 9x
    increase in poverty to occur in an area around here, but since the
    neighborhood boundaries do not correspond exactly with the census
    tracts, we cannot confidently confirm or refute the claim
    specifically for Holmesburg.

- Tioga-Nicetown (31): a predominantly black neighborhood located in
  Northwest Philadelphia: according to the article, the median income of
  this neighborhood dropped from 28,026 to 17,493, the largest drop in
  the city.

  - Analysis: in the census tract nearest to Tioga-Nicetown, a
    statistically significant increase in poverty can be observed. This
    corresponds with the 40 percent decrease in median income reported
    in the article. Areas bordering the neighborhood did not see
    statistically significant increases in poverty, however. The
    analysis in the article appears to be valid.

## Part 5: Conclusion

Philadelphia continues to suffer from high poverty rates, particularly
around minority-heavy areas in North Philly and West Philly. Income
inequality is particularly high here, with statistically significant
increases in wealth being observed between the 2010 and 2017 census.
However, many lower-income areas in North Philly are seeing
statistically significant decreases in poverty rates, which may suggest
a sign of hope for these communities.

Therefore, this analysis provides mixed evidence for the article's
claims. Changes in poverty were not uniform across Philadelphia, and
some areas in close proximity to each other experienced significant
increases in poverty while some experienced significant decreases. Many
other areas experienced no statistically significant changes at all. The
data in many of these census tracts was deemed either "somewhat
reliable" or "unreliable" based on its MOE, and a more detailed and
complete survey of this area is necessary to get fully accurate results.
Additionally, it would improve the accuracy of this analysis to conduct
the survey by neighborhood instead of by census tract; however, it would
be quite logistically challenging.
