---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Nathan Leroy"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

------------------------------------------------------------------------

## The question

Every number the American Community Survey publishes is an estimate
drawn from a sample, not a count. That means every number comes with a
margin of error — a statement about how wrong it might be.

In this analysis, I'm looking to see how population and the measurement
of reliability are related, using the number of college-educated
individuals in each county as an example.

------------------------------------------------------------------------

## Pulling the data

```{r}
my_state <- "CA"        # two-letter abbreviation of state
my_variable <- "B15003_022"     # variable code
```

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

This code displays the first few rows of our new county_data table.

```{r}
dim(county_data)
head(county_data)
```

::: callout-tip
Each of the 58 counties in CA are represented in this table, with GEOID,
name of the county, variable name, estimate, and margin of error forming
the 5 columns.
:::

------------------------------------------------------------------------

## How big is the margin of error?

A margin of error of 3,000 means something very different for an
estimate of 20,000 than for an estimate of 200,000. To compare across
places, we need it as a percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

Sierra County and Alpine County have by far the greatest MOE, followed
by Trinity, Modoc, and Colusa Counties.

------------------------------------------------------------------------

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 10 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

Using our reliability cutoffs, 28 counties fall into a "high confidence"
category, 13 into "moderate confidence", and 17 into "low confidence".
The percentiles were chosen somewhat arbitrarily, but they seem
reasonable for this assignment.

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the
same row. That means a wide pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

In this table, popE = estimate, and popM = MOE.

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

The following block compares the average population of counties in each
reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```

We can see here that population is clearly correlated with confidence in
deriving the population statistics; high-confidence counties have by far
the highest population, and low-confidence counties have by far the
lowest population numbers.

------------------------------------------------------------------------

## What this means for policy

Sierra County, located in a rural area in Northern California, has the
state's second-lowest population and scored the lowest in our
reliability estimate. The neighboring Alpine County, which has the
lowest population in the state, only scored slightly higher. Conversely,
Los Angeles County, which has the highest population in the state, had
the highest score for reliability.

This trend is related to how the ACS collects data; since it is based on
a sample percentage rather than a complete survey of every person in the
county, smaller counties have less people included in the study. With a
smaller sample size, chances are higher that there may be variations in
the study that aren't necessarily representative of the whole. On the
other hand, larger counties have larger samples that lead to more
reliable estimates, as any variation is generally smoothed out the large
number of observations.

For our variable (people with college educations), this could have the
potential impact of improper funding allocation to areas that already
have enough, as well as the diversion of education funds that could be
sorely needed in a smaller county. For example, if it is improperly
estimated that very people in Alpine County are college-educated (which
is highly correlated with their children not going to college), while
many people actually do want to go to college, the county could be
passed over when it's time to build a new university (like a CSU).
